---
title: "Final Project: Plastic Chemical Facilities and Maternal Health Explorer"
author: "Salem Yohannes and Megan Liu"
date: "2026-06-25"

#runtime: shiny 

params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

This project seeks to investigate whether there is a relationship between plastic chemical production and maternal health outcomes in the United States. 

Women in the US are at greater risk of experiencing adverse maternal health outcomes.[1] These disparate outcomes are a result of a variety of factors, including environmental exposures, systemic racism, inaccessibility to quality care, and policy inequities; and can persist regardless of protective factors such as socioeconomic status and education.[2] Exposure to chemicals involved in producing plastic is linked to maternal health effects such as decreased fertility, preterm birth, cesarean deliveries, and low birth weight.[3] These toxic chemicals can contaminate the environment during production, exposing communities near facilities to hazardous chemicals. Fenceline communities experience negative health outcomes from toxic exposures, including adverse maternal health impacts among women living in Louisiana’s Cancer Alley.[4] 

Examining whether the presence of plastic chemical production facilities is related to adverse maternal health outcomes in communities may improve understanding of disparities and identify areas where policy interventions may be needed. To investigate this relationship, an interactive dashboard was created to explore chemical facilities and maternal health by state. 

## Data

Maternal health outcomes were downloaded from the CDC’s National Vital Statistics System, which provides a range of birth-related statistics based on data from birth certificates.[5] Data on fertility rate, preterm births, cesarean deliveries, and low birth weight were downloaded, as these four outcomes are associated with exposure to plastic chemicals. Key variables included year, state names, and incidence rates of the four outcomes. The maternal health data was joined by year and state into one data frame in R. Additional data frames were created to prepare for the three dashboard visualizations. Line graph maternal health data was filtered for 2014 to 2024 to show trends over a 10-year period, with a new column calculating national averages. Scatterplot and bivariate map data was filtered for 2023, to match when chemical facility data was last reported. 

Chemical facility data was obtained from the most recent round of EPA Chemical Data Reporting in 2024, which contains information on chemical manufacturers and importers in the US.[6] Key variables included chemical names, chemical identification numbers, facility activity, and facility locations by state. The data was filtered to include 45 plastic chemicals (full list and chemical selection criteria linked [here](https://docs.google.com/spreadsheets/d/1SMKLifv7GmNUcRvL_HIJr1Ib90fFDOkrgmID7XN3OO8/edit?gid=0#gid=0)), the 50 states (territories were dropped), and manufacturing facilities (importing facilities, facilities under TSCA Confidential Business Information provisions, and missing values were dropped), resulting in 607 plastic chemical facilities included in the analysis.

The cleaned maternal health and chemical facility data were joined by state for the scatterplot. The combined data was then merged with state shapefiles using the {tigris} package for the bivariate choropleth map. 

```{r}
#| eval: false
# Load libraries and data
library(tidyverse)

# Load EPA Chemical Data Reporting (CDR) Data
cdr <- X2024_CDR_Manufacture_Import_Information
glimpse(cdr)

# Check how many rows of data
nrow(cdr) # 50,116 rows

# Drop CBI facilities, importing facilities, and missing values 
cdr <- cdr[cdr$`SITE NAICS ACTIVITY 1` != "CBI" & 
             !is.na(cdr$`SITE NAICS ACTIVITY 1`) &
             cdr$`SITE NAICS ACTIVITY 1` != "Import", ]

# Check how many rows of data
nrow(cdr) # 25,830, so dropped 24,286 rows total

# Load spreadsheet with specific plastic production chemicals and their CAS #s
plastic_cas_df <- Plastic_Production_Chemical_CAS
glimpse(plastic_cas_df)
  
# Create vector of all the CAS #s for easier filtering
plastic_cas <- plastic_cas_df$`CAS #` 
plastic_cas
  
# Filter cdr df down to just the plastic production chemicals identified
cdr_plastic <- cdr %>% 
  filter(`CHEMICAL ID` %in% plastic_cas) # 610 production facilities

# Upload maternal health data below

# Load libraries and data
#install.packages("dplyr")
library(dplyr)
library(readxl)

# Load LBW data by county from the National County Health Rankings Database 
lbw_county <- LBW_County_Data
glimpse(lbw_county)

# Load fertility rate and cesarean birth rate data by county from the CDC Wonder Natality Database 
fertility_cesarean_county <- Fertility_Rate_Cesarean_Delivery_Rate_County_Data
glimpse(fertility_cesarean_county)

# Load preterm birth data by county from the CDC Wonder Natality Database 
preterm_county <- Preterm_Birth_County_Data
glimpse(preterm_county)

#Load state data from CDC Stats of the State Data with maternal outcome data for 2005, 2014-2024
State_Data_2014_2024 <- read_excel("data/State Maternal Data 2005, 2014-2024.xlsx")
view(State_Data_2014_2024)

# Remove 2005 from state maternal outcome data to only have 2014-2024
# Add full state names as column
# Make year numeric 
state_data_2014_2024 <- State_Data_2014_2024 %>%
  filter(Year != 2005) %>%
  mutate(
    State_Name = state.name[match(State, state.abb)],
    Year = as.numeric(Year)
  ) %>%
  select(State_Name, State, Year, everything())
view(state_data_2014_2024)

# Keep only preterm birth data, which is birth before 37 weeks of gestational age
preterm_county <- Preterm_Birth_County_Data %>%
  filter(`OE Gestational Age` %in% c(
    "20 - 27 weeks",
    "28 - 31 weeks",
    "32 - 35 weeks",
    "36 weeks"
  )) 
glimpse(preterm_county)

# Add up all preterm birth data to get a total 
preterm_county <- preterm_county %>%
  group_by(State, County, Year) %>%
  summarise(
    preterm_births = sum(Births, na.rm = TRUE),
    .groups = "drop"
  )

# Make sure values are numeric because will be using total birth number from fertility_cesarean_county dataset in the preterm_county data
glimpse(preterm_county)
glimpse(fertility_cesarean_county)
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(Year = as.numeric(Year),
    `Total Number of Births` = as.numeric(`Total Number of Births`))

# Create a new column with the summarized preterm births 
# Calculate the total number of preterm births and the preterm birth rate
preterm_county <- preterm_county %>%
  left_join(
    fertility_cesarean_county %>% select(State, County, Year, `Total Number of Births`),
    by = c("State", "County", "Year")
  ) %>%
  mutate(
    preterm_birth_rate = preterm_births / `Total Number of Births`
  )
glimpse(preterm_county)

# Remove total state data in the rows of the LBW county data  
lbw_county <- lbw_county %>%
  filter(!`County Name` %in% state.name)
glimpse(lbw_county)
view(fertility_cesarean_county)

#Check datasets are all cleaned and ready to be used 
dim(preterm_county)
dim(fertility_cesarean_county)
dim(lbw_county)
dim(state_data_2014_2024)
glimpse(preterm_county)
glimpse(fertility_cesarean_county)
glimpse(lbw_county)
glimpse(state_data_2014_2024)

# Make characters numeric
# Change county name to be consistent with the chemical data
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(
    `Female Population` = as.numeric(`Female Population`),
    `Fertility Rate` = as.numeric(`Fertility Rate`))
preterm_county <- preterm_county %>%
  mutate(
    County = sub(",.*", "", County))
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(
    County = sub(",.*", "", County))

```   

```{r}
# Prepare chemical facility data
# Load necessary libraries and datasets
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(tidycensus)
library(leaflet)
library(dplyr)
library(shiny)
library(ggplot2)

options(tigris_use_cache = TRUE)

# Check for missing values in county name column
sum(is.na(cdr_plastic$`SITE COUNTY / PARISH`)) # 43 missing values

# Use the zipcodeR package which has zip codes and county names, to join county names by zip code to the 43 rows missing county name info
#install.packages("zipcodeR")
library(zipcodeR)

# Create df with zipcodes
zipcodeR_df <- zipcodeR::zip_code_db

# Join the county names to the cdr_plastic dataframe
zipcodeR_df <- zipcodeR_df %>% select(zipcode, county) # Select only relevant columns, so that the full dataframe doesn't get joined to cdr_plastic df

cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("SITE POSTAL CODE" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

names(cdr_plastic2) # Check that join worked, and all columns needed were added

# Join didn't work for zip codes that had the additional 4 digits. Update zip code column to only include the 5 first digits
 cdr_plastic <- cdr_plastic %>%  mutate(
  zip_code = substr(`SITE POSTAL CODE`, 1, 5))

# Re-join
cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("zip_code" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

# Check that join worked and there are no missing county names
sum(is.na(cdr_plastic$county_name)) # 0 missing county names

# Select for only necessary columns to make df more viewable
cdr_plastic3 <- cdr_plastic2 %>% 
  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, `SITE ADDRESS LINE1`, `SITE CITY`,
         `SITE COUNTY / PARISH`, `SITE STATE`, `SITE POSTAL CODE`, `SITE NAICS CODE 1`,
         `SITE NAICS ACTIVITY 1`, zipcodeR_county, zip_code)

# Now that all county names are included, add the ones from zipcodeR to fill the missing names in the original df
cdr_plastic4 <- cdr_plastic3 %>% mutate(
  county_name = coalesce(`SITE COUNTY / PARISH`, zipcodeR_county)) # Looking at the data, there are some wrong entries in the original data which is now under "county_name", so plan on actually using "zipcodeR_county" column for analysis, which has more accurate names.

# Rearrange columns so that it's easier to read
cdr_plastic5 <- cdr_plastic4 %>%  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, 
          `SITE ADDRESS LINE1`, `SITE CITY`, zipcodeR_county, `SITE STATE`, 
          zip_code, `SITE NAICS CODE 1`, `SITE NAICS ACTIVITY 1`)

# Standardize cases- county names all currently lower case
cdr_plastic5$zipcodeR_county <- toupper(cdr_plastic5$zipcodeR_county)

# Rename df
chemfacilities <- cdr_plastic5

# Download county shapefiles
counties_sf <- tigris::counties(cb = TRUE, year = 2023) # Chose to download 2023 since this is the most recent year of data in our CDR and maternal health data

# Standardize cases in shapefile df before merging
counties_sf$NAMELSAD <- toupper(counties_sf$NAMELSAD)
  
# Download state shapefiles so that state borders appear on map
states_sf <- tigris::states(cb = TRUE, year = 2023)

# Group by county and summarize number of facilities by county 
chemfacilities_counties <- chemfacilities %>% 
  group_by(zipcodeR_county, `SITE STATE`) %>% 
  summarise(count = n())

sum(chemfacilities_counties$count) # Check that all 610 facilities are included; true

# Join new count column to shapefile df (counties_sf) on county and state name
county_facilities_sf <- counties_sf %>% 
  left_join(chemfacilities_counties, by = c("NAMELSAD" = "zipcodeR_county", "STUSPS" = "SITE STATE"))

# Replace NA values in count column with 0
county_facilities_sf$count[is.na(county_facilities_sf$count)] <- 0 
sum(county_facilities_sf$count) # 607 total facilities, since the 3 in Puerto Rico were dropped
   
# Create univariate chloropleth just to visualize number of chemical facilities by county
tm_shape(county_facilities_sf) + # Read in county shapefile
  tm_polygons("count", palette = "brewer.reds", title = "Number of Chemical Facilities per County") +
  tm_shape(states_sf) + # Read in state shapefile
  tm_borders(col = "black", lwd = 2)

# Combine county maternal health outcomes with the shapefile df with facility counts

lbwcounty_forjoin <- lbw_county %>%  # Prepare the lbw_county df for join
  filter(Year == 2023) %>% 
  mutate(county_name = toupper(`County Name`)) %>% 
  select('State Abbreviation', county_name, 'Year', 'Low birthweight raw value')

pretermcounty_forjoin <- preterm_county %>% # Prepare the preterm_county df for join
  mutate(county_name = toupper(County)) %>% 
  filter(Year == 2023) %>% 
  select(State, county_name, preterm_birth_rate) 

fertility_cesarean_county_forjoin <- fertility_cesarean_county %>% 
  mutate(countyname = toupper(County)) %>% 
  filter(Year == 2023) %>% 
  select(State, countyname, `Fertility Rate`, `Cesarean Delivery Percent`)

# Join maternal health outcomes by state abbreviation and county names
county_facility_maternal_sf <- county_facilities_sf %>% 
  left_join(lbwcounty_forjoin, by = c("STUSPS" = "State Abbreviation", "NAMELSAD" = "county_name")) %>% 
  left_join(pretermcounty_forjoin, by = c("STATE_NAME" = "State", "NAMELSAD" = "county_name")) %>% 
  left_join(fertility_cesarean_county_forjoin, by = c("STATE_NAME" = "State", "NAMELSAD" = "countyname"))

sum(is.na(county_facility_maternal_sf$`Low birthweight raw value`)) # 123 counties with no lbw data
sum(is.na(county_facility_maternal_sf$preterm_birth_rate)) # 2558 counties with no preterm birth data
sum(is.na(county_facility_maternal_sf$`Fertility Rate`)) # 2558 counties with no fertility rate data
sum(is.na(county_facility_maternal_sf$`Cesarean Delivery Percent`)) # 2558 counties with no c-section rate data

# Export county_facility_maternal_sf into Excel, so can upload to Google as a public sheet for Shiny Assistant to reference to provide bivariate chloropleth map code
#install.packages("writexl")
library(writexl)
write_xlsx(county_facility_maternal_sf, "county_facility_maternal_sf.xlsx")


# Create new df with facility count by state, in case it's better visualization for map
chemfacilities_states <- chemfacilities %>% 
  group_by(`SITE STATE`) %>% 
  summarise(count = n())

# Join to state shapefile df
state_facilities_sf <- states_sf %>% 
  left_join(chemfacilities_states, by = c("STUSPS" = "SITE STATE"))

# Replace NA values in count column with 0
state_facilities_sf$count[is.na(state_facilities_sf$count)] <- 0 
sum(state_facilities_sf$count) # 610 states

# Drop territories
state_facilities_sf <- state_facilities_sf %>% 
  filter(!STUSPS %in% c("PR", "MP", "GU", "VI", "AS", "DC"))

# Create columns containing state averages grouped by state and year
state_maternalhealth_avgs <- State_Maternal_Outcome_Data %>% 
  group_by(Year, State) %>% 
  mutate(avg_cesarean = mean(`Cesarean Deliveries Percent`, na.rm = TRUE),
         avg_lbw = mean(`LBW Percent`, na.rm = TRUE),
         avg_fertility = mean(`Fertility Rate`, na.rm = TRUE), 
         avg_preterm = mean(`Preterm Births Percent`, na.rm = TRUE), 
         avg_births = mean(`Number of Births`, na.rm = TRUE))

# Join maternal health outcomes by state abbreviation 
state_facility_maternal_sf <- state_facilities_sf %>% 
  left_join(state_maternalhealth_avgs, by = c("STUSPS" = "State")) 
 
hist(chemfacilities_states$count, breaks = 50)
 
```

```{r}
# Aggregate number of chemical production facilities by state
state_facilities <- chemfacilities %>% 
  group_by(`SITE STATE`) %>% 
  summarise(count = n()) # 39 states have chemical facilities based on the chemicals included

# Check that the group_by() included all 610 facilities
sum(state_facilities$count) # = 610

# Create df with all 50 states in case we want to include for analysis
length(state.abb) # Use built in R vector, state.abb which has all 50 states (the CDR data had NA values and typos, otherwise would have extracted from there)

all_50 <- data.frame(state_abb = state.abb)

all50_state_facilities <- all_50 %>% 
  left_join(state_facilities, by = c(state_abb = "SITE STATE"))

all50_state_facilities$count[is.na(all50_state_facilities$count)] <- 0 # Replace NA values with 0 in case we want to explore this in the scatterplot

sum(all50_state_facilities$count) # = 607; 3 facilities now missing, because 3 facilities were in Puerto Rico which we will exclude from our analysis (?)

# Create new df with just 2023 rates for all 50 states, for scatterplot; and join facility counts
mhcf_2023_scatterdf <- state_maternalhealth_avgs %>% 
  left_join(all50_state_facilities, by = c("State" = "state_abb")) %>% 
  filter(Year %in% c(2023), State != "District of Columbia") %>% 
  select(avg_cesarean, avg_lbw, avg_fertility, avg_preterm, avg_births, count) 
   
sum(mhcf_2023_scatterdf$count) # Check that all facilities accounted for- true (all 607 present)

# Add column with state names for tooltip display
statenamesandabbvs <- data.frame(state_name = state.name, state_abb = state.abb)

# Join back to mhcf_2023_scatterdf
mhcf_2023_scatterdf <- mhcf_2023_scatterdf %>%  
  left_join(statenamesandabbvs, by = c("State" = "state_abb"))

# Create df option with only the 38 states that have facilities, in case we choose this route for visualization
## Update: we will not use this dataframe, and will include all 50 states, but leaving here in case
mhcf_2023_scatterdf_only38 <- mhcf_2023_scatterdf %>% 
  filter(count != 0)

# Create scatterplot
mhcf_scatterplot <- mhcf_2023_scatterdf %>% # mh = maternal health, cf = chemical facilities
  ggplot(aes(x = count, y = avg_cesarean)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Number of Chemical Facilities and Maternal Health Outcomes", 
       x = "Number of Chemical Facilities per State", 
       y = "Average Rate of Cesarean Deliveries") +
  theme_minimal()
mhcf_scatterplot

# Notes: for Shiny, we can alter title label and y axis label to reflect which outcome is selected for

```

```{r}
# your code here
# mhcf_2023_scatterdf # Use this for plotting

# your code below 
# state_data_2014_2024 # Use this for plotting
# view(state_data_2014_2024)
# view(all50_state_facilities)
 
# creating national averages for each maternal health outcome 
  
ntl_maternal_avgs <- state_data_2014_2024 %>%
  filter(State != "District of Columbia") %>%   #filter out DC
  group_by(Year) %>%
    summarise(
    ntl_cesarean = mean(`Cesarean Deliveries Percent`, na.rm = TRUE),
    ntl_lbw = mean(`Low Birth Weight Percent`, na.rm = TRUE),
    ntl_fertility = mean(`Fertility Rate per 1,000 women`, na.rm = TRUE),
    ntl_preterm = mean(`Preterm Birth Percent`, na.rm = TRUE),
    .groups = "drop"
  )

  view(ntl_maternal_avgs)       
        
```

## Shiny App: Plastic Chemical Facilities and Maternal Health Explorer

The Plastic Chemical Facilities and Maternal Health Explorer is an interactive dashboard that investigates the relationship between chemical facilities and maternal health. Users can select for a single maternal health outcome at a time (fertility rate, preterm birth, cesarean deliveries, or low birth weight), which then produces three visualizations for exploration. To contextualize maternal health in the US, the line graph first describes trends over time, and compares state-level maternal health data to the national average. Next, the scatterplot shows the relationship between the frequency of facilities and maternal health by state. Lastly, the map provides an interactive visualization of the geographic relationship between chemical facilities and maternal health.

[Click here to open the interactive Shiny app](https://n4s5zp-megan-liu.shinyapps.io/shiny-app/)

## Interpretation

Key findings from the visualizations revealed state variation in both maternal health outcomes and the presence of chemical facilities. The line graph showed that while national maternal health trends remained relatively stable over time, individual states often deviated from the national average. Comparisons with the national average identified states that consistently performed above or below national trends across the time series. 

States with higher counts of chemical facilities often appeared to have higher rates of adverse maternal health outcomes, but this pattern was not consistent across the US. Texas had the highest number of chemical facilities (182), and Louisiana had the second highest (114), with Louisiana consistently standing out with both a high amount of facilities and worse maternal health outcomes. Higher rates of adverse maternal health outcomes were commonly observed in central and southern states, across varying levels of chemical facility counts. These visualizations highlight states that experience compounded impacts of disproportionate exposure to chemical production facilities and adverse maternal health outcomes, which can help identify areas for further research, policy, and public health interventions.  

## Limitations

This project had a few limitations. The visualizations compared state-level data which was useful for higher level comparisons, but does not allow for a more granular analysis. County-level maternal health data would have provided more specific, community-level insights, however comprehensive maternal health outcome data was not available at the county-level. 

The chemical facility data was also limited to a small amount of plastic production chemicals, rather than a comprehensive list of the thousands of chemicals involved in plastic production. It is likely that there are many other plastic chemical facilities that were not included in this exploration. A more in-depth literature review could result in a more comprehensive list of plastic chemical facilities being included in the analysis.

Overall, this dashboard was an exploratory analysis, so was able to identify states with higher counts of plastic chemical facilities and adverse maternal health rates, but did not establish association. With additional time and stronger data, future work could incorporate county-level geographical visualizations with a more comprehensive list of chemicals and facilities, as well as perform statistical association tests. 

## Conclusion

This investigation aims to provide users, especially public health stakeholders and policymakers, with an understanding of maternal health status in the US, and whether the presence of plastic chemical production facilities in communities has any connection to adverse maternal health outcomes. By comparing state trends to national averages over time, and identifying states that have greater exposures to plastic pollution and worse maternal health outcomes, users can easily identify areas that may be disproportionately impacted. 

## References

1. Munira Gunja et al., Insights into the U.S. Maternal Mortality Crisis: An International Comparison (Commonwealth Fund, June 2024). https://doi.org/10.26099/cthn-st75 
2. Hailu, E. M., Maddali, S. R., Snowden, J. M., Carmichael, S. L., & Mujahid, M. S. (2022). Structural racism and adverse maternal health outcomes: A systematic review. Health & place, 78, 102923. https://doi.org/10.1016/j.healthplace.2022.102923 
3. Qian, Y., Shao, H., Ying, X., Huang, W., & Hua, Y. (2020). The endocrine disruption of prenatal phthalate exposure in mother and offspring. Frontiers in Public Health, 8, 366. https://doi.org/10.3389/fpubh.2020.00366
4. Juhasz, A. (2026). “We’re dying here.” In Human Rights Watch. https://www.hrw.org/report/2024/01/25/were-dying-here/the-fight-for-life-in-a-louisiana-fossil-fuel-sacrifice-zone#5194
5. NVSS - National Vital Statistics System Homepage. (n.d.). https://www.cdc.gov/nchs/nvss/index.htm
6. Access CDR Data | US EPA. (2026, May 19). US EPA. https://www.epa.gov/chemical-data-reporting/access-cdr-data
