---
title: "Final Project: Plastic Production Chemicals and Maternal Health Outcomes"
author: "Salem Yohannes , Megan Liu"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

_Describe your data source(s), including where it came from, the key variables, and any pre-processing steps._

```{r}
#| eval: false
# Load libraries and data
library(tidyverse)

# Load EPA Chemical Data Reporting (CDR) Data
cdr <- X2024_CDR_Manufacture_Import_Information
glimpse(cdr)

# Check how many rows of data
nrow(cdr) ## 50,116 rows

# Drop CBI facilities, importing facilities, and missing values 
cdr <- cdr[cdr$`SITE NAICS ACTIVITY 1` != "CBI" & 
             !is.na(cdr$`SITE NAICS ACTIVITY 1`) &
             cdr$`SITE NAICS ACTIVITY 1` != "Import", ]

# Check how many rows of data
nrow(cdr) ## 25,830, so dropped 24,286 rows total

# Load spreadsheet with specific plastic production chemicals and their CAS #s
plastic_cas_df <- Plastic_Production_Chemical_CAS
glimpse(plastic_cas_df)
  
# Create vector of all the CAS #s for easier filtering
plastic_cas <- plastic_cas_df$`CAS #` 
plastic_cas
  
# Filter cdr df down to just the plastic production chemicals identified
cdr_plastic <- cdr %>% 
  filter(`CHEMICAL ID` %in% plastic_cas) # 610 production facilities

#Load additional data below
```   

## Visualizatiob 1: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 2: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 3: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## OR Shiny app: [Title Here]

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

