---
title: "Final Project: Plastic Production Chemicals and Maternal Health Outcomes"
author: "Salem Yohannes , Megan Liu"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

_Describe your data source(s), including where it came from, the key variables, and any pre-processing steps._

```{r}
#| eval: false
# Load libraries and data
library(tidyverse)

# Load EPA Chemical Data Reporting (CDR) Data
cdr <- X2024_CDR_Manufacture_Import_Information
glimpse(cdr)

# Check how many rows of data
nrow(cdr) # 50,116 rows

# Drop CBI facilities, importing facilities, and missing values 
cdr <- cdr[cdr$`SITE NAICS ACTIVITY 1` != "CBI" & 
             !is.na(cdr$`SITE NAICS ACTIVITY 1`) &
             cdr$`SITE NAICS ACTIVITY 1` != "Import", ]

# Check how many rows of data
nrow(cdr) # 25,830, so dropped 24,286 rows total

# Load spreadsheet with specific plastic production chemicals and their CAS #s
plastic_cas_df <- Plastic_Production_Chemical_CAS
glimpse(plastic_cas_df)
  
# Create vector of all the CAS #s for easier filtering
plastic_cas <- plastic_cas_df$`CAS #` 
plastic_cas
  
# Filter cdr df down to just the plastic production chemicals identified
cdr_plastic <- cdr %>% 
  filter(`CHEMICAL ID` %in% plastic_cas) # 610 production facilities


-----------
# Upload maternal health data below

# Load libraries and data
install.packages("dplyr")
library(dplyr)
library(readxl)

# Load LBW data by county from the National County Health Rankings Database 
lbw_county <- LBW_County_Data
glimpse(lbw_county)

# Load fertility rate and cesarean birth rate data by county from the CDC Wonder Natality Database 
fertility_cesarean_county <- Fertility_Rate_Cesarean_Delivery_Rate_County_Data
glimpse(fertility_cesarean_county)

# Load preterm birth data by county from the CDC Wonder Natality Database 
preterm_county <- Preterm_Birth_County_Data
glimpse(preterm_county)

#Load state data from CDC Stats of the State Data with maternal outcome data for 2005, 2014-2024
State_Data_2014_2024 <- read_excel("data/State Maternal Data 2005, 2014-2024.xlsx")
view(State_Data_2014_2024)

# Remove 2005 from state maternal outcome data to only have 2014-2024
# Add full state names as column
# Make year numeric 
state_data_2014_2024 <- State_Data_2014_2024 %>%
  filter(Year != 2005) %>%
  mutate(
    State_Name = state.name[match(State, state.abb)],
    Year = as.numeric(Year)
  ) %>%
  select(State_Name, State, Year, everything())
view(state_data_2014_2024)

# Keep only preterm birth data, which is birth before 37 weeks of gestational age
preterm_county <- Preterm_Birth_County_Data %>%
  filter(`OE Gestational Age` %in% c(
    "20 - 27 weeks",
    "28 - 31 weeks",
    "32 - 35 weeks",
    "36 weeks"
  )) 
glimpse(preterm_county)

# Add up all preterm birth data to get a total 
preterm_county <- preterm_county %>%
  group_by(State, County, Year) %>%
  summarise(
    preterm_births = sum(Births, na.rm = TRUE),
    .groups = "drop"
  )

# Make sure values are numeric because will be using total birth number from fertility_cesarean_county dataset in the preterm_county data
glimpse(preterm_county)
glimpse(fertility_cesarean_county)
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(Year = as.numeric(Year),
    `Total Number of Births` = as.numeric(`Total Number of Births`))

# Create a new column with the summarized preterm births 
# Calculate the total number of preterm births and the preterm birth rate
preterm_county <- preterm_county %>%
  left_join(
    fertility_cesarean_county %>% select(State, County, Year, `Total Number of Births`),
    by = c("State", "County", "Year")
  ) %>%
  mutate(
    preterm_birth_rate = preterm_births / `Total Number of Births`
  )
glimpse(preterm_county)

# Remove total state data in the rows of the LBW county data  
lbw_county <- lbw_county %>%
  filter(!`County Name` %in% state.name)
glimpse(lbw_county)
view(fertility_cesarean_county)

#Check datasets are all cleaned and ready to be used 
dim(preterm_county)
dim(fertility_cesarean_county)
dim(lbw_county)
dim(state_data_2014_2024)
glimpse(preterm_county)
glimpse(fertility_cesarean_county)
glimpse(lbw_county)
glimpse(state_data_2014_2024)

# Make characters numeric
# Change county name to be consistent with the chemical data
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(
    `Female Population` = as.numeric(`Female Population`),
    `Fertility Rate` = as.numeric(`Fertility Rate`))
preterm_county <- preterm_county %>%
  mutate(
    County = sub(",.*", "", County))
fertility_cesarean_county <- fertility_cesarean_county %>%
  mutate(
    County = sub(",.*", "", County))

```   

## Visualization 1: Bivariate Chloropleth Map

_Describe what this plot shows, and why it's important._

```{r}
# Prepare chemical facility data
# Load necessary libraries and datasets
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(tidycensus)
library(leaflet)
library(dplyr)
library(shiny)
library(ggplot2)

options(tigris_use_cache = TRUE)

# Check for missing values in county name column
sum(is.na(cdr_plastic$`SITE COUNTY / PARISH`)) # 43 missing values

# Use the zipcodeR package which has zip codes and county names, to join county names by zip code to the 43 rows missing county name info
install.packages("zipcodeR")
library(zipcodeR)

# Create df with zipcodes
zipcodeR_df <- zipcodeR::zip_code_db

# Join the county names to the cdr_plastic dataframe
zipcodeR_df <- zipcodeR_df %>% select(zipcode, county) # Select only relevant columns, so that the full dataframe doesn't get joined to cdr_plastic df

cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("SITE POSTAL CODE" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

names(cdr_plastic2) # Check that join worked, and all columns needed were added

# Join didn't work for zip codes that had the additional 4 digits. Update zip code column to only include the 5 first digits
cdr_plastic <- cdr_plastic %>%  mutate(
  zip_code = substr(`SITE POSTAL CODE`, 1, 5))

# Re-join
cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("zip_code" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

# Check that join worked and there are no missing county names
sum(is.na(cdr_plastic$county_name)) # 0 missing county names

# Select for only necessary columns to make df more viewable
cdr_plastic3 <- cdr_plastic2 %>% 
  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, `SITE ADDRESS LINE1`, `SITE CITY`,
         `SITE COUNTY / PARISH`, `SITE STATE`, `SITE POSTAL CODE`, `SITE NAICS CODE 1`,
         `SITE NAICS ACTIVITY 1`, zipcodeR_county, zip_code)

# Now that all county names are included, add the ones from zipcodeR to fill the missing names in the original df
cdr_plastic4 <- cdr_plastic3 %>% mutate(
  county_name = coalesce(`SITE COUNTY / PARISH`, zipcodeR_county)) # Looking at the data, there are some wrong entries in the original data which is now under "county_name", so plan on actually using "zipcodeR_county" column for analysis, which has more accurate names.

# Rearrange columns so that it's easier to read
cdr_plastic5 <- cdr_plastic4 %>%  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, 
          `SITE ADDRESS LINE1`, `SITE CITY`, zipcodeR_county, `SITE STATE`, 
          zip_code, `SITE NAICS CODE 1`, `SITE NAICS ACTIVITY 1`)

# Standardize cases- county names all currently lower case
cdr_plastic5$zipcodeR_county <- toupper(cdr_plastic5$zipcodeR_county)

# Rename df
chemfacilities <- cdr_plastic5

# Download county shapefiles
counties_sf <- tigris::counties(cb = TRUE, year = 2023) # Chose to download 2023 since this is the most recent year of data in our CDR and maternal health data

# Standardize cases in shapefile df before merging
counties_sf$NAMELSAD <- toupper(counties_sf$NAMELSAD)
  
# Download state shapefiles so that state borders appear on map
states_sf <- tigris::states(cb = TRUE, year = 2023)

# Group by county and summarize number of facilities by county 
chemfacilities_counties <- chemfacilities %>% 
  group_by(zipcodeR_county, `SITE STATE`) %>% 
  summarise(count = n())

sum(chemfacilities_counties$count) # Check that all 610 facilities are included; true

# Join new count column to shapefile df (counties_sf) on county and state name
county_facilities_sf <- counties_sf %>% 
  left_join(chemfacilities_counties, by = c("NAMELSAD" = "zipcodeR_county", "STUSPS" = "SITE STATE"))

# Replace NA values in count column with 0
county_facilities_sf$count[is.na(county_facilities_sf$count)] <- 0 
sum(county_facilities_sf$count) # 607 total facilities, since the 3 in Puerto Rico were dropped
   
# Create univariate chloropleth just to visualize number of chemical facilities by county
tm_shape(county_facilities_sf) + # Read in county shapefile
  tm_polygons("count", palette = "brewer.reds", title = "Number of Chemical Facilities per County") +
  tm_shape(states_sf) + # Read in state shapefile
  tm_borders(col = "black", lwd = 2)

# Create interactive leaflet map


 
```

## Visualization 2: Scatterplot

_Describe what this plot shows, and why it's important._

```{r}
# Aggregate number of chemical production facilities by state
state_facilities <- chemfacilities %>% 
  group_by(`SITE STATE`) %>% 
  summarise(count = n()) # 39 states have chemical facilities based on the chemicals included

# Check that the group_by() included all 610 facilities
sum(state_facilities$count) # = 610

# Create df with all 50 states in case we want to include for analysis
length(state.abb) # Use built in R vector, state.abb which has all 50 states (the CDR data had NA values and typos, otherwise would have extracted from there)

all_50 <- data.frame(state_abb = state.abb)

all50_state_facilities <- all_50 %>% 
  left_join(state_facilities, by = c(state_abb = "SITE STATE"))

all50_state_facilities$count[is.na(all50_state_facilities$count)] <- 0 # Replace NA values with 0 in case we want to explore this in the scatterplot

sum(all50_state_facilities$count) # = 607; 3 facilities now missing, because 3 facilities were in Puerto Rico which we will exclude from our analysis (?)

# Create columns containing state averages grouped by state and year
state_maternalhealth_avgs <- State_Maternal_Outcome_Data %>% 
  group_by(Year, State) %>% 
  mutate(avg_cesarean = mean(`Cesarean Deliveries Percent`, na.rm = TRUE),
         avg_lbw = mean(`LBW Percent`, na.rm = TRUE),
         avg_fertility = mean(`Fertility Rate`, na.rm = TRUE), 
         avg_preterm = mean(`Preterm Births Percent`, na.rm = TRUE), 
         avg_births = mean(`Number of Births`, na.rm = TRUE))

# Create new df with just 2023 rates for all 50 states, for scatterplot; and join facility counts
mhcf_2023_scatterdf <- state_maternalhealth_avgs %>% 
  left_join(all50_state_facilities, by = c("State" = "state_abb")) %>% 
  filter(Year %in% c(2023), State != "District of Columbia") %>% 
  select(avg_cesarean, avg_lbw, avg_fertility, avg_preterm, avg_births) %>% 
  left_join(all50_state_facilities, by = c("State" = "state_abb"))

sum(mhcf_2023_scatterdf$count) # Check that all facilities accounted for- true (all 607 present)

# Create df option with only the 38 states that have facliities, in case we choose this route for visualization
mhcf_2023_scatterdf_only38 <- mhcf_2023_scatterdf %>% 
  filter(count != 0)

# Create scatterplot
mhcf_scatterplot <- mhcf_2023_scatterdf %>% # mh = maternal health, cf = chemical facilities
  ggplot(aes(x = count, y = avg_cesarean)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Number of Chemical Facilities and Maternal Health Outcomes", 
       x = "Number of Chemical Facilities per State", 
       y = "Average Rate of Cesarean Deliveries") +
  theme_minimal()
mhcf_scatterplot


# Notes: for Shiny, we can alter title label and y axis label to reflect which outcome is selected for

```

## Visualization 3: Faceted Line Graph

_Describe what this plot shows, and why it's important._

```{r}
# your code here
mhcf_2023_scatterdf # Use this for plotting



-------
# your code below 
state_data_2014_2024 # Use this for plotting
  
  
```

## OR Shiny app: Plastic Chemical Production Facilities and Maternal Health in the US

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

