---
title: "Final Project: Plastic Production Chemicals and Maternal Health Outcomes"
author: "Salem Yohannes , Megan Liu"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

_Describe your data source(s), including where it came from, the key variables, and any pre-processing steps._

```{r}
#| eval: false
# Load libraries and data
library(tidyverse)

# Load EPA Chemical Data Reporting (CDR) Data
cdr <- X2024_CDR_Manufacture_Import_Information
glimpse(cdr)

# Check how many rows of data
nrow(cdr) # 50,116 rows

# Drop CBI facilities, importing facilities, and missing values 
cdr <- cdr[cdr$`SITE NAICS ACTIVITY 1` != "CBI" & 
             !is.na(cdr$`SITE NAICS ACTIVITY 1`) &
             cdr$`SITE NAICS ACTIVITY 1` != "Import", ]

# Check how many rows of data
nrow(cdr) # 25,830, so dropped 24,286 rows total

# Load spreadsheet with specific plastic production chemicals and their CAS #s
plastic_cas_df <- Plastic_Production_Chemical_CAS
glimpse(plastic_cas_df)
  
# Create vector of all the CAS #s for easier filtering
plastic_cas <- plastic_cas_df$`CAS #` 
plastic_cas
  
# Filter cdr df down to just the plastic production chemicals identified
cdr_plastic <- cdr %>% 
  filter(`CHEMICAL ID` %in% plastic_cas) # 610 production facilities

# Upload maternal health data below

```   

## Visualizatiob 1: Bivariate Chloropleth Map

_Describe what this plot shows, and why it's important._

```{r}
# Prepare chemical facility data
# Load necessary libraries and datasets
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(tidycensus)
library(leaflet)

# Check for missing values in county name column
sum(is.na(cdr_plastic$`SITE COUNTY / PARISH`)) # 43 missing values

# Use the zipcodeR package which has zip codes and county names, to join county names by zip code to the 43 rows missing county name info
install.packages("zipcodeR")
library(zipcodeR)

# Create df with zipcodes
zipcodeR_df <- zipcodeR::zip_code_db

# Join the county names to the cdr_plastic dataframe
zipcodeR_df <- zipcodeR_df %>% select(zipcode, county) # Select only relevant columns, so that the full dataframe doesn't get joined to cdr_plastic df

cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("SITE POSTAL CODE" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

names(cdr_plastic2) # Check that join worked, and all columns needed were added

# Join didn't work for zip codes that had the additional 4 digits. Update zip code column to only include the 5 first digits
cdr_plastic <- cdr_plastic %>%  mutate(
  zip_code = substr(`SITE POSTAL CODE`, 1, 5))

# Re-join
cdr_plastic2 <- cdr_plastic %>% 
  left_join(zipcodeR_df, by = c("zip_code" = "zipcode")) %>% 
  mutate(zipcodeR_county = county)

# Check that join worked and there are no missing county names
sum(is.na(cdr_plastic$county_name)) # 0 missing county names

# Select for only necessary columns to make df more viewable
cdr_plastic3 <- cdr_plastic2 %>% 
  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, `SITE ADDRESS LINE1`, `SITE CITY`,
         `SITE COUNTY / PARISH`, `SITE STATE`, `SITE POSTAL CODE`, `SITE NAICS CODE 1`,
         `SITE NAICS ACTIVITY 1`, zipcodeR_county)

# Now that all county names are included, add the ones from zipcodeR to fill the missing names in the original df
cdr_plastic4 <- cdr_plastic3 %>% mutate(
  county_name = coalesce(`SITE COUNTY / PARISH`, zipcodeR_county)) # Looking at the data, there are some wrong entries in the original data which is now under "county_name", so plan on actually using "zipcodeR_county" column for analysis, which has more accurate names.

# Rearrange columns so that it's easier to read
cdr_plastic5 <- cdr_plastic4 %>%  select(`CHEMICAL NAME`, `CHEMICAL ID`, `SITE NAME`, 
          `SITE ADDRESS LINE1`, `SITE CITY`, zipcodeR_county, `SITE STATE`, 
          `SITE POSTAL CODE`, `SITE NAICS CODE 1`, `SITE NAICS ACTIVITY 1`)

# Standardize cases- county names all currently lower case
cdr_plastic5$zipcodeR_county <- toupper(cdr_plastic5$zipcodeR_county)

toupper(counties_mw$NAME) 

# Rename df
chemfacilities <- cdr_plastic5

# Prepare spatial data

```

## Visualization 2: Scatterplot

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 3: Faceted Line Graph

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## OR Shiny app: Plastic Chemical Production Facilities and Maternal Health in the US

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

